### 2013-07-27 doc/taxize_withcode_plos.Rnw
incorporated changes from karthik and naim, fixes #170, should be close to submit

% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{taxize - taxonomic search and retrieval in R}
}
% Insert Author names, affiliations and corresponding author email.
\\
Scott Chamberlain$^{1}$ and 
Eduard Sz\"{o}cs$^{2}$
\\
\bf{1} Biology Department, Simon Fraser University, Canada
\\
\bf{2} Institute for Environmental Sciences, University Koblenz-Landau, Fortstr. 7, 76829 Landau, Germany
\\
$\ast$ E-mail: myrmecocystus@gmail.com
\end{flushleft}

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}

@@

The R language is a widely used language by biologists, and now has over 5,000 packages on the Comprehensive R Archive Network (CRAN) and more than 2,500 packages on other repositories to extend R. R is great for manipulating, visualizing and fitting statistical models to data. \cite{gentleman_bioconductor:_2004} gives a detailed discussion of advantages of R in computational biology. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Therefore there is a need to get data from the web directly into R. Increasingly, data is available from the web via application programming interfaces (API). These allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further advantage of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Moreover data can be accessed from every computer, whereas for example an Excel file can only be opened in a few programs.

The goal of taxize, an R package in development, is to make many use cases having to do with retrieving and resolving taxonomic names easy and reproducible. In taxize, we have written a suite of R functions that interact with many taxonomic data sources via their web APIs (Table~\ref{tab:a}). The interface to each function is usually a simple list of species names, just as a user would do with a web API. Therefore, we hope moving from a web to R interface for taxonomic names will be relatively seamless (if one is already nominally familiar with R).

Here, we justify the need for taxize, discuss our data sources, and run through a suite of use cases to demonstrate the variety of ways that users can interact with taxize.

@@

All of these data repositories provide ways to search for taxonomic names and resolve them in some cases. However, scientists ideally need a tool that can be used programmatically, and thereby facilitate reproducible research. The goal of taxize is to make it easy to create reproducible and easy to use workflows for searching for taxonomic names, resolving them, getting higher taxonomic names, and other tasks related to research dealing with species.

\section*{Data sources and package details}
taxize uses many data sources (Table~\ref{tab:a}), and more can easily be added. There are two common tasks provided by the data sources: name search and name resolution. Other functionality in taxize includes retrieving a classification tree for a species, or retrieving child taxa of a focal taxon. One of the data sources (Phylomatic) returns phylogenies, while another (NCBI) returns genetic sequence data. However, there are other R packages that are focused solely on sequence data, such as rsnps \cite{chamberlain2013}, rentrez \cite{winter2013}, BoSSA \cite{lefeuvre2010}, and ape \cite{paradis2004}, so taxize will not venture deeply into these other domains.

Some of the data sources taxize interacts with require authentication. That is, in addition to the search terms the user provides (e.g., \emph{Homo sapiens}), the data provider requires an alphanumeric identification key so that they can better manage their servers, collect analytics, and shut down users that abuse the API. The services that do require an API key in taxize are: Encyclopedia of Life (EOL) \cite{eol}, the Universal Biological Indexer and Organizer (uBio) \cite{ubio}, Tropicos \cite{tropicos}, and Plantminer \cite{carvalho2010plantminer}. One can easily obtain API keys by visiting the website of each service (see Table~\ref{tab:a} for links to each site). There are two typical ways of using API keys. First, you can pass in your API key in a function call (e.g., \emph{ubio\_namebank(srchName='Ursus americanus', key='your\_alphanumeric\_key')}). Second, you can store your key in the .Rprofile file, which is a common place to store settings. We recommend the second option as it simplifies function calls as taxize detects the stored keys.

One available data source in taxize is The Plant List \cite{theplantlist}. The connection in taxize is done via the \emph{taxonstand} package \cite{Cayuela_2012} that solely interacts with that The Plant List. We provide a few convenience functions that wrap taxonstand into taxize.

@@

New data sources can be added; we plan to add the following sources: Wikispecies and The Tree of Life. A connection to \url{freshwaterecology.info} \cite{freshwaterecology} (a database with autecological characteristics, ecological preferences and biological traits as well as distribution patterns of more than 12.000 European freshwater organisms belonging to fish, macro-invertebrates, macrophytes, diatoms and phytoplankton) will be finished when their new API will be released.
In addition, the authors may be contacted for further suggestions of data sources to be added.

@@

There are a variety of use cases for which taxize is ideally suited. We discuss here several ideal use cases for taxize. Moreover a complete reproducible workflow from a species list to a phylogeny and a distribution map can be found in the supplement.

@@

<<resolvenames, eval=TRUE, cache=TRUE>>=
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"), returndf = TRUE)
temp[ , -c(1,4)]
@

@@

<<tnrs, eval=TRUE,cache=TRUE>>=
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
  	"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")
tnrs(query = mynames)[ , -c(5:7)]
@

@@

@@

In another example, you can pass in a long list of taxonomic names:

@@

It turns out that there are XX different unique records found. However, this doesn't mean that there are XX different genes found as the API does not provide metadata to classify genes. However, at the end of the example, we showed that you can use regular expressions (e.g., via \emph{grep}) to search for the gene of interest.

@@

We see that the two Juncus species are aggregated to Juncaceae and their abundances are summed up. If a taxon is one lower taxonomic resolution then the querried or the taxon is not found in the database (like in the example taxon \emph{xxx}) then these are not aggregated and returned as is.

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
The taxize package is part of the rOpenSci project \url{http://ropensci.org/}. We thank X, Y, and Z for comments on previous versions of this manuscript. SAC is supported by CANPOLIN of Canada, grant number XXXXXX. We thank all API maintainers for their work making their databases open to the public.

@@

\section*{Figure Legends}
\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{figure/phylomaticphylogeny.pdf}
\end{center}
\caption{
{\bf A Phylogeny for the three species.}
}
\label{fig:phylomatic}
\end{figure}

>>>>

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
\doublespacing

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{taxize: taxonomic search and retrieval in R}
}
% Insert Author names, affiliations and corresponding author email.
\\
Scott Chamberlain$^{1}$ and 
Eduard Sz\"{o}cs$^{2}$
\\
\bf{1} Biology Department, Simon Fraser University, Burnaby, BC, Canada
\\
\bf{2} Institute for Environmental Sciences, University Koblenz-Landau, Fortstr. 7, 76829 Landau, Germany
\\
$\ast$ E-mail: myrmecocystus@gmail.com
\end{flushleft}

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
% \section*{Author Summary}

@@

The R language is a widely used language by biologists, and now has over 5,000 packages on the Comprehensive R Archive Network (CRAN) to extend R. R is great for manipulating, visualizing and fitting statistical models to data. \cite{gentleman_bioconductor:_2004} gives a detailed discussion of advantages of R in computational biology. Getting data from the web will be increasingly common as more and more data gets moved to the cloud. Therefore there is a need to get data from the web directly into R. Increasingly, data is available from the web via application programming interfaces (API). These allow computers to talk to one another using code that is not human readable, but is machine readable. Web APIs often define a number of methods that allow users to search for a species name, or retrieve the synonyms for a species name, for example. A further advantage of APIs is that they are language agnostic, meaning that data can be consumed in almost any computing context, allowing users to interact with the web API without having to know the details of the code. Moreover data can be accessed from every computer, whereas for example an Excel file can only be opened in a few programs.

The goal of taxize, an R package in development, is to make many use cases having to do with retrieving and resolving taxonomic names easy and reproducible. In taxize, we have written a suite of R functions that interact with many taxonomic data sources via their web APIs (Table~\ref{tab:a}). The interface to each function is usually a simple list of species names, just as a user would do when interacting with a website. Therefore, we hope moving from a web to R interface for taxonomic names will be relatively seamless (if one is already nominally familiar with R).

Here, we justify the need for programmatic taxonomic resolution tools like taxize, discuss our data sources, and run through a suite of use cases to demonstrate the variety of ways that users can use taxize.

@@

All of these data repositories provide ways to search for taxonomic names and resolve them in some cases. However, scientists ideally need a tool that is free and can be used programmatically,  thereby facilitating reproducible research. The goal of taxize is to make it easy to create reproducible and easy to use workflows for searching for taxonomic names, resolving them, getting higher taxonomic names, and other tasks related to research dealing with species.

\section*{Data sources and package details}
taxize uses many data sources (Table~\ref{tab:a}), and more can easily be added. There are two common tasks provided by the data sources: name search and name resolution. Other functionality in taxize includes retrieving a classification tree for a species, or retrieving child taxa of a focal taxon. One of the data sources (Phylomatic) returns phylogenies, while another (NCBI) returns genetic sequence data. However, there are other R packages that are focused solely on sequence data, such as rsnps \cite{chamberlain2013}, rentrez \cite{winter2013}, BoSSA \cite{lefeuvre2010}, and ape \cite{paradis2004}, so taxize does not venture deeply into these other domains.

Some of the data sources taxize interacts with require authentication. That is, in addition to the search terms the user provides (e.g., \emph{Homo sapiens}), the data provider requires an alphanumeric identification key so that they can better manage their servers, collect analytics, and shut down users that abuse the API. The services that require an API key in taxize are: Encyclopedia of Life (EOL) \cite{eol}, the Universal Biological Indexer and Organizer (uBio) \cite{ubio}, Tropicos \cite{tropicos}, and Plantminer \cite{carvalho2010plantminer}. One can easily obtain API keys by visiting the website of each service (see Table~\ref{tab:a} for links to each site). There are two typical ways of using API keys. First, you can pass in your API key in a function call (e.g., \emph{ubio\_namebank(srchName='Ursus americanus', key='your\_alphanumeric\_key')}). Second, you can store your key in the .Rprofile file, which is a common place to store settings. We recommend the second option as it simplifies function calls as taxize detects the stored keys.

One available data source in taxize is The Plant List \cite{theplantlist}. The connection in taxize is done via the \emph{taxonstand} package \cite{Cayuela_2012} that solely interacts with that The Plant List. We provide a few convenience functions that wrap \emph{taxonstand} into taxize.

@@

New data sources can be added; for example, we plan to add the following sources: Wikispecies and The Tree of Life. A connection to \url{freshwaterecology.info} \cite{freshwaterecology} (a database with autecological characteristics, ecological preferences and biological traits as well as distribution patterns of more than 12,000 European freshwater organisms belonging to fish, macro-invertebrates, macrophytes, diatoms and phytoplankton) will be finished when their new API will be released. In addition, the authors may be contacted for further suggestions of data sources to be added.

@@

There are a variety of use cases for which taxize is ideally suited. We discuss here several ideal use cases for taxize. Moreover a complete reproducible workflow from a species list to a phylogeny and a distribution map can be found in Appendix A.

@@

<<resolvenames, eval=TRUE, cache=TRUE>>=
temp <- gnr_resolve(names = c("Helianthos annus", "Homo saapiens"))
temp[ , -c(1,4)]
@

@@

<<tnrs, eval=TRUE, cache=TRUE>>=
mynames <- c("Helianthus annuus", "Pinus contort", "Poa anua", "Abis magnifica",
  	"Rosa california", "Festuca arundinace", "Sorbus occidentalos","Madia sateva")
tnrs(query = mynames)[ , -c(5:7)]
@

@@

Another common use case is when there are many synonyms for a species. In this example, we have three synonyms of the currently accepted name for a species.

<<synonyms, eval=TRUE, cache=TRUE, message=FALSE, warning=FALSE, comment=NA>>=
mynames <- c("Helianthus annuus ssp. jaegeri", "Helianthus annuus ssp. lenticularis", "Helianthus annuus ssp. texanus", "Helianthus annuus var. lenticularis", "Helianthus annuus var. macrocarpus", "Helianthus annuus var. texanus")
tsn <- get_tsn(mynames)
ldply(tsn, itis_acceptname)
@

@@

In another example, you can pass in a long character vector of taxonomic names:

@@

It turns out that there are 430 different unique records found. However, this doesn't mean that there are 430 different genes found as the API does not provide metadata to classify genes. However, you can use regular expressions (e.g., \emph{grep}) to search for the gene of interest.

@@

We see that the two Juncus species are aggregated to Juncaceae and their abundances are summed up. If a taxon is one lower taxonomic resolution then the querried or the taxon is not found in the database then these are not aggregated and returned as is.

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
The taxize package is part of the rOpenSci project \url{http://ropensci.org/}. We thank Carl Boettiger, Karthik Ram, Owen Jones, Naim Matasci, and Ralf Sch\"{a}fer for comments on previous versions of this manuscript. We thank all API maintainers for their work making their databases open to the public.

@@

\section*{Figure Legends}
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{figure/phylomaticphylogeny.pdf}
\end{center}
\caption{
{\bf A Phylogeny for the three species.}
}
\label{fig:phylomatic}
\end{figure}
